---
title: "Getting Started with emulate: Target Trial Emulation in R"
author: "Timothy P. Copeland"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{Getting Started with emulate}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.width = 7,
  fig.height = 5
)
```

## Introduction

This vignette walks through a complete target trial emulation analysis using
the emulate package. We will analyze a bundled example dataset and produce two
analyses:

1. **Intention-to-Treat (ITT)**: What is the effect of being *assigned* to
   treatment, regardless of actual adherence?
2. **Per-Protocol (PP)**: What is the effect of *adhering* to the assigned
   treatment strategy?

By the end, you will understand every step of the emulate pipeline and be able
to adapt it to your own data.

## Loading the Package and Data

```{r load-data}
library(emulate)

# Load the bundled trial_example.csv dataset
# This ships with emulate in inst/extdata/
trial_data <- read.csv(
  system.file("extdata", "trial_example.csv", package = "emulate")
)
```

## Understanding the Data

Let us examine the structure of the dataset:

```{r str-data}
str(trial_data)
```

```{r summary-data}
summary(trial_data)
```

```{r dim-data}
cat("Rows:", nrow(trial_data), "\n")
cat("Unique patients:", length(unique(trial_data$id)), "\n")
cat("Period range:", min(trial_data$period), "-", max(trial_data$period), "\n")
cat("Events:", sum(trial_data$outcome == 1), "\n")
cat("Eligible observations:", sum(trial_data$eligible == 1), "\n")
```

The dataset contains **503 patients** observed over approximately 100 time
periods, resulting in **48,400 person-period observations**. This is
person-period format: each row represents one patient at one time period.

### Variable definitions

| Variable | Type | Description |
|----------|------|-------------|
| `id` | Integer | Unique patient identifier |
| `eligible` | Binary (0/1) | Whether patient can enter a new trial at this period |
| `period` | Integer | Calendar time period |
| `outcome` | Binary (0/1) | Terminal event indicator |
| `treatment` | Binary (0/1) | Current treatment status |
| `catvarA`, `catvarB`, `catvarC` | Integer | Categorical covariates |
| `nvarA`, `nvarB`, `nvarC` | Numeric | Continuous covariates |

## Step 0: Define the Protocol

Before touching the data, we should define the target trial we are emulating.
The `emulate_protocol()` function creates the 7-component protocol table from
Hernan and Robins:

```{r protocol, eval = FALSE}
protocol <- emulate_protocol(
  eligibility    = "Adults meeting clinical criteria, not currently on treatment,
                    no prior outcome event",
  treatment      = "Strategy A: Initiate treatment at the start of each eligible
                    period. Strategy B: Do not initiate treatment.",
  assignment     = "Patients are assigned to treatment strategies at each period
                    when they meet eligibility criteria.",
  followup_start = "Start of the period when eligibility is first met for each
                    emulated trial.",
  outcome        = "First occurrence of the binary outcome event during follow-up.
                    Maximum follow-up: 8 periods.",
  causal_contrast = "Intention-to-treat effect and per-protocol effect of
                     treatment initiation vs. no treatment.",
  analysis       = "Sequential trial emulation with pooled logistic regression.
                    PP analysis uses stabilized IPTW with truncation at 1st/99th
                    percentiles. G-formula marginal predictions with 100 MC
                    samples for confidence intervals.",
  format = "display"
)
```

This step produces no statistical output but forces you to articulate the
study design. It can also be exported to Excel or LaTeX for inclusion in
manuscripts.

## Analysis 1: Intention-to-Treat

The ITT analysis estimates the effect of treatment *assignment*, ignoring
whether patients actually adhere. This is the simplest and most robust
analysis.

### Step 1: Prepare the data

```{r itt-prepare}
itt <- emulate_prepare(
  trial_data,
  id        = "id",
  period    = "period",
  treatment = "treatment",
  outcome   = "outcome",
  eligible  = "eligible",
  covariates = c("catvarA", "catvarB", "catvarC",
                 "nvarA", "nvarB", "nvarC"),
  estimand  = "ITT"
)
```

`emulate_prepare()` maps the variable names, validates the data structure, and
creates a `emulate` object. The summary tells us how many individuals, periods,
and events are in the dataset.

### Step 2: Validate the data

```{r itt-validate}
itt <- emulate_validate(itt)
```

The 10 checks examine data quality: no duplicate rows, no gaps in period
sequences, outcome is terminal, treatment variation among eligible, and so
on. All checks should pass or produce only notes and warnings, not errors.

### Step 3: Expand into sequential trials

```{r itt-expand}
itt <- emulate_expand(itt, maxfollowup = 8)
```

This is the core of the sequential trials method. For ITT, each eligible
individual at each trial period is enrolled and followed for up to 8 periods.
No cloning or artificial censoring occurs under ITT -- individuals keep their
observed treatment and are followed as-is.

The expansion ratio tells you how much the dataset grew. The expanded dataset
contains one row per individual per trial per follow-up period.

### Step 4: Fit the outcome model

```{r itt-fit}
itt <- emulate_fit(
  itt,
  outcome_cov = c("catvarA", "catvarB", "catvarC",
                   "nvarA", "nvarB", "nvarC"),
  followup_spec      = "quadratic",
  trial_period_spec  = "quadratic"
)
```

The pooled logistic regression models the probability of the outcome at each
follow-up period, adjusting for treatment arm, follow-up time (quadratic),
trial period (quadratic), and covariates. Standard errors are clustered by
patient ID.

The treatment effect is reported as a log-odds ratio and odds ratio. A
negative log-odds ratio indicates that treatment is protective.

### Step 5: Predict cumulative incidence

```{r itt-predict}
itt <- emulate_predict(
  itt,
  times      = 0:8,
  type       = "cum_inc",
  difference = TRUE,
  samples    = 100,
  seed       = 12345
)
```

The G-formula computes marginal (population-averaged) cumulative incidence
curves for each arm. Monte Carlo simulation provides confidence intervals by
drawing from the sampling distribution of the model coefficients.

With `difference = TRUE`, risk differences (treated minus control) are also
computed at each time point.

### Step 6: Plot cumulative incidence

```{r itt-plot, eval = FALSE}
emulate_plot(itt, type = "cumhaz", title = "ITT: Cumulative Incidence")
```

The plot shows cumulative incidence curves for the treatment and control arms
with 95% confidence bands. Separation between the curves indicates a
treatment effect.

### Step 7: Report results

```{r itt-report}
emulate_report(itt, format = "display", eform = TRUE, predictions = TRUE)
```

The report summarizes the analysis: estimand, model type, sample sizes, and
the coefficient table. With `eform = TRUE`, coefficients are exponentiated
to odds ratios.

To export to Excel:

```{r itt-excel, eval = FALSE}
emulate_report(itt, format = "excel", export = "itt_results.xlsx",
           eform = TRUE, predictions = TRUE)
```

### Interpretation

Look at the treatment arm coefficient (the odds ratio for `_emulate_arm`). An
OR < 1 indicates treatment is protective; an OR > 1 indicates treatment is
harmful. The confidence interval and p-value tell you about statistical
significance.

The cumulative incidence predictions show the absolute risk over time in each
arm. The risk difference at the final time point gives you the absolute
treatment effect.

## Analysis 2: Per-Protocol

The PP analysis estimates the effect of treatment *adherence* -- what would
happen if everyone followed their assigned strategy. This requires modeling
treatment switching with inverse probability weights.

### Step 1: Prepare the data (PP)

```{r pp-prepare}
pp <- emulate_prepare(
  trial_data,
  id        = "id",
  period    = "period",
  treatment = "treatment",
  outcome   = "outcome",
  eligible  = "eligible",
  covariates = c("catvarA", "catvarB", "catvarC",
                 "nvarA", "nvarB", "nvarC"),
  estimand  = "PP"
)
```

### Step 2: Validate

```{r pp-validate}
pp <- emulate_validate(pp)
```

Note the treatment consistency check: for PP, the validator reports the
number of treatment discontinuations. These are the events that will trigger
artificial censoring during expansion.

### Step 3: Expand into sequential trials (PP)

```{r pp-expand}
pp <- emulate_expand(pp, maxfollowup = 8)
```

Under PP, each eligible individual is **cloned** into two copies: one
assigned to the treatment arm (arm = 1) and one assigned to the control arm
(arm = 0). Each clone is then followed forward and **artificially censored**
when their observed treatment deviates from the assigned strategy:

- Treatment arm clone: censored when the individual *stops* treatment
- Control arm clone: censored when the individual *starts* treatment

This is the "clone-censor-weight" framework. The expansion ratio is roughly
twice that of ITT because of the cloning.

### Step 4: Compute inverse probability weights

```{r pp-weight}
pp <- emulate_weight(
  pp,
  switch_d_cov = c("catvarA", "catvarB", "catvarC",
                    "nvarA", "nvarB", "nvarC"),
  switch_n_cov = c("catvarA", "catvarB"),
  truncate     = c(1, 99),
  stabilized   = TRUE
)
```

This is the key step that distinguishes PP from ITT. The weight models
estimate the probability of treatment switching and create stabilized
inverse probability weights to correct for the informative censoring
introduced by the cloning.

The **denominator model** includes all time-varying covariates -- it models
the probability of the observed treatment given the full history. The
**numerator model** includes only baseline covariates for stabilization.

The weight summary shows mean, SD, min, max, and effective sample size (ESS).
For well-behaved weights:
- Mean should be close to 1 (for stabilized weights)
- SD should not be too large
- ESS should be a reasonable fraction of the sample size

### Step 5: Fit the outcome model (PP)

```{r pp-fit}
pp <- emulate_fit(
  pp,
  outcome_cov = c("catvarA", "catvarB", "catvarC",
                   "nvarA", "nvarB", "nvarC"),
  followup_spec      = "quadratic",
  trial_period_spec  = "quadratic"
)
```

The PP outcome model is the same as ITT but uses the inverse probability
weights. The weighted regression corrects for the bias introduced by
artificial censoring.

### Step 6: Predict cumulative incidence (PP)

```{r pp-predict}
pp <- emulate_predict(
  pp,
  times      = 0:8,
  type       = "cum_inc",
  difference = TRUE,
  samples    = 100,
  seed       = 12345
)
```

### Step 7: Plot results

```{r pp-plot-ci, eval = FALSE}
emulate_plot(pp, type = "cumhaz", title = "PP: Cumulative Incidence")
```

```{r pp-plot-weights, eval = FALSE}
emulate_plot(pp, type = "weights", title = "PP: IP Weight Distribution")
```

The weight distribution plot shows the density of weights by arm. Ideally,
weights are concentrated near 1 with thin tails. Heavy tails or extreme
values suggest the weight model may need refinement or tighter truncation.

### Step 8: Diagnose weight quality

```{r pp-diagnose}
pp <- emulate_diagnose(
  pp,
  balance_covariates = c("catvarA", "catvarB", "catvarC",
                         "nvarA", "nvarB", "nvarC")
)
```

The diagnostics report:

- **Weight distribution**: Check that the mean is near 1 and the range is
  not extreme.
- **Effective sample size (ESS)**: A large drop from the actual sample size
  to the ESS indicates extreme weights.
- **Standardized mean differences (SMD)**: The unweighted and weighted SMD
  for each covariate. After weighting, all SMDs should be below 0.1
  (conventional threshold).

If balance is poor, consider:
- Adding more covariates to the switch model
- Using a different functional form
- Tightening the truncation percentiles

To visualize the balance:

```{r pp-plot-balance, eval = FALSE}
emulate_plot(pp, type = "balance", title = "PP: Covariate Balance (Love Plot)")
```

### Step 9: Report results

```{r pp-report}
emulate_report(pp, format = "display", eform = TRUE, predictions = TRUE)
```

### Interpretation

Compare the PP treatment effect to the ITT effect. The PP effect is often
larger in magnitude because it estimates the effect under full adherence
rather than the diluted intent-to-treat effect. However, the PP estimate
relies on the correctness of the weight model -- if the weight model is
misspecified, the PP estimate may be biased.

## Comparing ITT and PP

```{r compare}
cat("ITT treatment log-OR:", round(itt$model$b_treat, 4), "\n")
cat("PP  treatment log-OR:", round(pp$model$b_treat, 4), "\n")
cat("\n")
cat("ITT treatment OR:", round(exp(itt$model$b_treat), 4), "\n")
cat("PP  treatment OR:", round(exp(pp$model$b_treat), 4), "\n")
```

Key points for comparison:

- The **ITT estimate** is robust to model misspecification for treatment
  switching but is diluted by non-adherence. It answers: "What happens when
  we assign patients to treatment?"

- The **PP estimate** attempts to estimate the effect under full adherence
  but depends on correct weight modeling. It answers: "What would happen if
  everyone followed the protocol?"

- If the ITT and PP estimates are similar, treatment switching is either rare
  or does not substantially bias the ITT estimate.

- If the PP estimate is substantially larger in magnitude, non-adherence is
  attenuating the ITT effect.

- If the PP estimate is in the opposite direction from ITT, this is a red
  flag and may indicate weight model misspecification.

## Next Steps

### Cox proportional hazards model

```{r cox-example, eval = FALSE}
# You can use a Cox PH model instead of pooled logistic regression
obj <- emulate_fit(obj, model = "cox",
               outcome_cov = c("x1", "x2"),
               trial_period_spec = "quadratic")
# Note: emulate_predict() currently only supports the logistic model.
# After a Cox fit, predictions are not available.
```

### Sensitivity analyses

Robust analyses should assess sensitivity to key modeling choices:

1. **Truncation levels**: Try `c(1, 99)`, `c(5, 95)`, and `c(10, 90)` to
   see how weight trimming affects estimates.

2. **Time specifications**: Compare `"linear"`, `"quadratic"`, `"cubic"`,
   and `"ns(3)"` for both follow-up and trial period.

3. **Follow-up length**: Vary `maxfollowup` to see how the estimate changes
   with longer or shorter follow-up.

4. **Grace periods**: If treatment switching is gradual, a grace period of
   1-3 periods may be more realistic than strict censoring at the first
   deviation.

```{r sensitivity, eval = FALSE}
# Grace period analysis
pp_grace <- emulate_expand(pp, maxfollowup = 8, grace = 2)
```

### As-Treated analysis

For absorbing treatments (once started, never stopped), the As-Treated
estimand may be appropriate:

```{r at-example, eval = FALSE}
at <- emulate_prepare(trial_data, ..., estimand = "AT")
at <- emulate_expand(at, maxfollowup = 8)
at <- emulate_weight(at, switch_d_cov = c("x1", "x2"), truncate = c(1, 99))
at <- emulate_fit(at, outcome_cov = c("x1", "x2"))
```

For absorbing treatment, AT and PP produce identical results.

## References

- Hernan MA, Robins JM. Using big data to emulate a target trial when a
  randomized trial is not available. *American Journal of Epidemiology*.
  2016;183(8):758-764.

- Hernan MA, Robins JM. *Causal Inference: What If*. Boca Raton: Chapman &
  Hall/CRC, 2020.

- Maringe C, Benitez Majano S, et al. TrialEmulation: An R package for
  target trial emulation. *arXiv*. 2024;2402.12083.

- Maringe C, et al. Reflection on modern methods: trial emulation in the
  presence of immortal-time bias. *International Journal of Epidemiology*.
  2020;49(5):1719-1729.
